### 2013-10-01 manuscripts/dpp-msbayes/dpp-msbayes.tex
Working on Mark's edits of dpp ms.

{\sffamily
    \noindent\textbf{ABSTRACT} \\
    \noindent To understand the processes that generate biodiversity, it is
    important to account for large-scale processes that effect the evolutionary
    history of groups of co-distributed populations of organisms.
    Such events predict temporally clustered divergences times, a pattern
    that can be estimated from genetic data from co-distributed species.
    We introduce a new approximate-Bayesian model for comparative
    phylogeographical model-choice that estimates the temporal distribution of
    divergences across taxa from multi-locus DNA sequence data.
    Our model is an extension of the model implemented in \msb.
    By reparameterizing the model, and using more flexible priors on divergence
    models and nuisance parameters, we improve the robustness, accuracy, and power
    of the method for estimating the posterior probabilities of models
    of divergence across taxa.
    Our results demonstrate that the bias of the \msb model toward inferring
    models of clustered divergences is caused by a combination of (1) uniform
    priors on nuisance parameters hindering the marginal likelihoods of models
    with more divergence time parameters, and (2) a prior on divergence models
    that disfavors models with intermediate numbers of divergence time
    parameters.

@@

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of Evolutionary Biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can effect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history
across effected populations of organisms, e.g., temporally clustered
divergences across co-distributed species.
At recent timescales, such events should leave a signature in the genetic
variation within and among the effected lineages.
Thus, being able to accurately estimate models of shared evolutionary
histories across co-distributed taxa would provide an important
tool for better understanding how regional and global biogeographic
processes affect biodiversity.

@@

We assume an investigator is interested in inferring the distribution
of divergence times among \npairs{} pairs of populations.
For each pair $i$, \popSampleSize{i}{} genome copies have been sampled, with
\popSampleSize{1}{i} copies sampled from population 1, and \popSampleSize{2}{i}
sampled from population 2.
From these genomes, let \nloci{i} be the number of DNA sequence loci collected
for population pair $i$, and \nlociTotal be the total number of unique loci
sampled across the \npairs{} pairs of populations.
We use \alignment{i}{j} to represent the multiple sequence alignment of
locus $j$ for population pair $i$.
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
is the full dataset,
i.e., a vector of sequence alignments for all pairs and loci.
We let \geneTree{i}{j} represent the gene tree upon which \alignment{i}{j}
evolved according to fixed \hky substitution model parameters \hkyModel{i}{j}.
The investigtor must specify the parameters of all
$\hkyModelVector = (\hkyModel{1}{1}, \ldots,
\hkyModel{\npairs{}}{\nloci{\npairs{}}})$
substitution models by which the alignments evolved along the
$\geneTreeVector = (\geneTree{1}{1}, \ldots,
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
gene trees.
Furthermore, the investigator must specify a vector of fixed contants
$\ploidyScalarVector = (\ploidyScalar{1}{1}, \ldots,
\ploidyScalar{\npairs{}}{\nloci{\npairs{}}})$
that scale the population size parameters for known differences in ploidy among
loci and/or differences in generation times among population pairs.
Lastly, the investigator must also specify a vector of fixed constants
$\mutationRateScalarConstantVector = (\mutationRateScalarConstant{1}{1},
\ldots, \mutationRateScalarConstant{\npairs{}}{\nloci{\npairs{}}})$
that scale the population size parameters for known differences in
mutation rates among loci and/or among taxa.

@@

\subsubsection*{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The likelihood terms of Equation \ref{eq:fullModelCompact} can
be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$,
is the probability of the sequence alignment of locus $j$ for population pair
$i$ given the gene tree and \hky substitution model parameters \citep[i.e., the
``Felsenstein likelhood'';][]{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams contains the following demographic parameters:
The mutation-scaled sizes of the ancestral, \ancestralTheta{}, and descendant
populations, \descendantTheta{1}{} and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persists during bottlenecks that begin immmediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \citep{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

@@

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence time parameters are in coalescent units relative to the size of
a constant reference population, which we denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    \myTheta{C} = E(\descendantTheta{}{}).
    \label{eq:thetaC}
\end{equation}
Given the size of the descendant populations are \iid as
$\descendantTheta{}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
The \divTime{} parameters are then in units $\myTheta{C}/\mutationRate$
generations, which we denote as \globalcoalunit generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which we
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{
        \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
        {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, and thus comparable, the relative mutation rates among the pairs of
populations are assumed to be fixed and known according the user-provided
values in \mutationRateScalarConstantVector.
\end{linenomath}

@@

The second prior we implement is based on the Dirichlet-process, which is a
stochastic process that groups elements into an unknown number of discrete
parameter classes \citep{Ferguson1973,Antoniak1974}.
The process is controlled by the concentration parameter \concentrationParam,
which determines how clustered the process will be.
We use the Dirichlet process to place a prior over all possible ordered
partitions of \npairs{} elements, which we denote as $\divTimeIndexVector \sim
\priorDPP{}$.
We take a hierarchical approach and use a second-order prior probability
distribution (i.e., hyperprior) for \concentrationParam.
More specifically, we use a gamma-distributed prior $\concentrationParam =
Gamma(\gammaShape{\concentrationParam}, \gammaScale{\concentrationParam})$,
where \gammaShape{\concentrationParam} and \gammaScale{\concentrationParam} are
specified by the user.

\begin{linenomath}
This provides a great deal of flexibility for specifying the prior uncertainty
regarding divergence models.
The \concentrationParam determines the prior probability that any two
pairs of populations $i$ and $j$ will be assigned to the same divergence time
parameter
\begin{equation}
    p(\divTimeIndex{i} = \divTimeIndex{j}) = \frac{1}{1 + \concentrationParam},
    \label{eq:dppPriorSameClass}
\end{equation}
and also the prior probability of the number of divergence time parameters
\begin{equation}
    p(\divTimeNum \given \concentrationParam, \npairs{}) = 
    \frac{\stirlingFirst{\npairs{}}{\divTimeNum} \concentrationParam^{\divTimeNum}}
    {\prod_{i=1}^{\npairs{}}(\concentrationParam + i - 1)},
    \label{eq:dppPriorNumClasses}
\end{equation}
where \stirlingFirst{\cdot}{\cdot} are the unsigned Stirling numbers of the
first kind.
Equations \ref{eq:dppPriorSameClass} and \ref{eq:dppPriorNumClasses} show that
smaller values of \concentrationParam will favor fewer divergence time
parameters, and thus more clustered models of divergence, whereas larger values
will favor more divergence time parameters, and thus less clustered models of
divergence.
\end{linenomath}

\subsection*{Differences between our model and the original \msb model}
\subsubsection*{The prior on divergence models}
One of the key differences between our model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the only prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
We denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an odd U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, we place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
We introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides the user with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

Another distinction between our model and that of \msb, is the prior on the
sizes of the descendant populations of each pair.
As described by \citet{Oaks2012}, \msb uses the joint prior
\begin{equation}
    \descendantTheta{1}{}, \descendantTheta{2}{} \sim
    Dirichlet(1,1) \times 2 \times U(\uniformMin{\myTheta{}},
    \uniformMax{\descendantTheta{}{}}),
    \label{eq:jointThetaPrior}
\end{equation}
such that the user-specified uniform prior on descendant population
size is a prior on the \emph{mean} size of the two descendant
populations of each pair.
Under our model, the sizes of the descendant populations of each
pair are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$
and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$.
This relaxes the assumption that the sizes of the two descendant populations
are interdependent and negatively correlated.

>>>>

{\sffamily
    \noindent\textbf{ABSTRACT} \\
    \noindent To understand the processes that generate biodiversity, it is
    important to account for large-scale processes that affect the evolutionary
    history of groups of co-distributed populations of organisms.
    Such events predict temporally clustered divergences times, a pattern
    that can be estimated from genetic data from co-distributed species.
    We introduce a new approximate-Bayesian model for comparative
    phylogeographical model-choice that estimates the temporal distribution of
    divergences across taxa from multi-locus DNA sequence data.
    Our model is an extension of the model implemented in \msb.
    By reparameterizing the model, and using more flexible priors on divergence
    models and nuisance parameters, we improve the robustness, accuracy, and power
    of the method for estimating the posterior probabilities of models
    of divergence across taxa.
    Our results demonstrate that the bias of the \msb model toward inferring
    models of clustered divergences is caused by a combination of (1) uniform
    priors on nuisance parameters hindering the marginal likelihoods of models
    with more divergence time parameters, and (2) a prior on divergence models
    that disfavors models with intermediate numbers of divergence time
    parameters.

@@

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of evolutionary biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can affect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history
across effected populations of organisms, e.g., temporally clustered
divergences across co-distributed species.
At recent timescales, such events should leave a signature in the genetic
variation within and among the affected lineages.
Thus, being able to accurately estimate models of shared evolutionary
histories across co-distributed taxa would provide an important
tool for better understanding how regional and global biogeographic
processes affect biodiversity.

@@

We assume an investigator is interested in inferring the distribution
of divergence times among \npairs{} pairs of populations.
For each pair $i$, \popSampleSize{i}{} genome copies have been sampled, with
\popSampleSize{1}{i} copies sampled from population 1, and \popSampleSize{2}{i}
sampled from population 2.
From these genomes, let \nloci{i} be the number of DNA sequence loci collected
for population pair $i$, and \nlociTotal be the total number of unique loci
sampled across the \npairs{} pairs of populations.
We use \alignment{i}{j} to represent the multiple sequence alignment of
locus $j$ for population pair $i$.
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
is the full dataset,
i.e., a vector of sequence alignments for all pairs and loci.
We let \geneTree{i}{j} represent the gene tree upon which \alignment{i}{j}
evolved according to fixed \hky substitution model parameters \hkyModel{i}{j}.
The investigator must specify the parameters of all
$\hkyModelVector = (\hkyModel{1}{1}, \ldots,
\hkyModel{\npairs{}}{\nloci{\npairs{}}})$
substitution models by which the alignments evolved along the
$\geneTreeVector = (\geneTree{1}{1}, \ldots,
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
gene trees.
Furthermore, the investigator must specify a vector of fixed contants
$\ploidyScalarVector = (\ploidyScalar{1}{1}, \ldots,
\ploidyScalar{\npairs{}}{\nloci{\npairs{}}})$
that scale the population size parameters for known differences in ploidy among
loci and/or differences in generation times among population pairs.
Lastly, the investigator must also specify a vector of fixed constants
$\mutationRateScalarConstantVector = (\mutationRateScalarConstant{1}{1},
\ldots, \mutationRateScalarConstant{\npairs{}}{\nloci{\npairs{}}})$
that scale the population size parameters for known differences in
mutation rates among loci and/or among taxa.

@@

\subsubsection*{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The likelihood terms of Equation \ref{eq:fullModelCompact} can
be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$,
is the probability of the sequence alignment of locus $j$ for population pair
$i$ given the gene tree and \hky substitution model parameters \citep[i.e., the
``Felsenstein likelihood'';][]{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams{} contains the following demographic parameters:
The mutation-scaled sizes of the ancestral, \ancestralTheta{}, and descendant
populations, \descendantTheta{1}{} and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persist during bottlenecks that begin immediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \citep{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

@@

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence time parameters are in coalescent units relative to the size of
a constant reference population, which we denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}).
    \label{eq:thetaC}
\end{equation}
Given the size of the descendant populations are \iid as
$\descendantTheta{}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
The \divTime{} parameters are in units $\myTheta{C}/\mutationRate$
generations, which we denote as \globalcoalunit generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which we
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{
        \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
        {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be fixed and known according the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

@@

The second prior we implement is based on the Dirichlet-process, which is a
stochastic process that groups elements into an unknown number of discrete
parameter classes \citep{Ferguson1973,Antoniak1974}.
The process is controlled by the concentration parameter \concentrationParam,
which determines how clustered the process will be.
We use the Dirichlet process to place a prior over all possible ordered
partitions of \npairs{} elements, which we denote as $\divTimeIndexVector \sim
\priorDPP{}$.
We take a hierarchical approach and use a prior probability
distribution (i.e., hyperprior) for \concentrationParam.
More specifically, we use a gamma-distributed prior $\concentrationParam =
Gamma(\gammaShape{\concentrationParam}, \gammaScale{\concentrationParam})$,
where \gammaShape{\concentrationParam} and \gammaScale{\concentrationParam} are
specified by the user.

\begin{linenomath}
This provides a great deal of flexibility for specifying the prior uncertainty
regarding divergence models.
The \concentrationParam determines the prior probability that any two
pairs of populations $i$ and $j$ will be assigned to the same divergence time
parameter
\begin{equation}
    p(\divTimeIndex{i} = \divTimeIndex{j}) = \frac{1}{1 + \concentrationParam},
    \label{eq:dppPriorSameClass}
\end{equation}
and also the prior probability of the number of divergence time parameters
\begin{equation}
    p(\divTimeNum \boldsymbol{\mid} \concentrationParam, \npairs{}) = 
    \frac{\stirlingFirst{\npairs{}}{\divTimeNum} \concentrationParam^{\divTimeNum}}
    {\prod_{i=1}^{\npairs{}}(\concentrationParam + i - 1)},
    \label{eq:dppPriorNumClasses}
\end{equation}
where \stirlingFirst{\cdot}{\cdot} are the unsigned Stirling numbers of the
first kind.
Equations \ref{eq:dppPriorSameClass} and \ref{eq:dppPriorNumClasses} show that
smaller values of \concentrationParam will favor fewer divergence time
parameters, and thus more clustered models of divergence, whereas larger values
will favor more divergence time parameters, and thus less clustered models of
divergence.
\end{linenomath}

\subsection*{Differences between our model and the original \msb model}
\subsubsection*{The prior on divergence models}
One of the key differences between our model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the only prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
We denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, we place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
We introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides the user with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

Another distinction between our model and that of \msb, is the prior on the
sizes of the descendant populations of each pair.
As described by \citet{Oaks2012}, \msb uses the joint prior
\begin{equation}
    \descendantTheta{1}{}, \descendantTheta{2}{} \sim
    Beta(1,1) \times 2 \times U(\uniformMin{\myTheta{}},
    \uniformMax{\descendantTheta{}{}}),
    \label{eq:jointThetaPrior}
\end{equation}
such that the user-specified uniform prior on descendant population
size is a prior on the \emph{mean} size of the two descendant
populations of each pair.
Under our model, the sizes of the descendant populations of each
pair are \iid as
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$
and
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$.
This relaxes the assumption that the sizes of the two descendant populations
are interdependent and negatively correlated.
